---
title: "Homework Assignment 3: Analyzing Drug Use"
author: "Ryan Avery and Team Member2"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: pdf_document
---


```{r setup, echo=FALSE}
library(knitr)
# set global chunk options: images will be 7x5 inches
knitr::opts_chunk$set(fig.width=7, fig.height=5)
options(digits = 4)


## indents are for indenting r code as formatted text
## They may need to be adjusted depending on your OS
# if your output looks odd, increase or decrease indent
indent1 = '    '
indent2 = '        '
indent3 = '            '
```

```{r load, echo=FALSE}
library(tidyverse)
drug_use <- read_csv('drug.csv',
col_names = c('ID','Age','Gender','Education','Country','Ethnicity',
'Nscore','Escore','Oscore','Ascore','Cscore','Impulsive',
'SS','Alcohol','Amphet','Amyl','Benzos','Caff','Cannabis',
'Choc','Coke','Crack','Ecstasy','Heroin','Ketamine',
'Legalh','LSD','Meth','Mushrooms','Nicotine','Semer','VSA'))
```

1. Logistic Regression
```{r mutate, indent=indent2}
drug_use <- drug_use %>% mutate_at(as.ordered, .vars=vars(Alcohol:VSA))
drug_use <- drug_use %>%
mutate(Gender = factor(Gender, labels=c("Male", "Female"))) %>%
mutate(Ethnicity = factor(Ethnicity, labels=c("Black", "Asian", "White",
"Mixed:White/Black", "Other",
"Mixed:White/Asian",
"Mixed:Black/Asian"))) %>%
mutate(Country = factor(Country, labels=c("Australia", "Canada", "New Zealand",
"Other", "Ireland", "UK", "USA")))
```
    a) 
```{r cannabis, indent=indent2}
cannabis_levels <- c("No","Yes")
drug_use <- drug_use %>% 
  mutate(recent_cannabis_use = factor(ifelse(Cannabis>='CL3',"Yes","No")),cannabis_levels) 

```
    b)  
```{r split, out.width='0.4\\linewidth', fig.show='hold', indent=indent1}
drug_use_subset <- drug_use %>% select(Age:SS, recent_cannabis_use)
set.seed(1)
train.indices = sample(1:nrow(drug_use_subset), 1500)
drug_use_train=drug_use_subset[train.indices,]
drug_use_test=drug_use_subset[-train.indices,]
dim(drug_use_train)
dim(drug_use_test)
```

  c)
```{r logreg, out.width='50%', fig.show='hold', indent=indent1}
logistic.fit = glm(recent_cannabis_use ~ ., data=drug_use_train, family=binomial(link="logit"))
summary(logistic.fit)
```
  d)
```{r}

probit.fit = glm(recent_cannabis_use ~ ., data=drug_use_train, family=binomial(link="probit"))

plot(x = fitted(logistic.fit), y=fitted(probit.fit), pch=19, cex=0.2)
abline(a=0, b=1, col="red")
title("Probit vs Logistic Model")


```

```{r}

cloglog.fit = glm(recent_cannabis_use ~ ., data=drug_use_train, family=binomial(link="cloglog"))

plot(x = fitted(logistic.fit), y=fitted(cloglog.fit), pch=19, cex=0.2)
abline(a=0, b=1, col="red")
title("cloglog vs Logistic Model")

```

The probit model yields predicted values that are most similar to the logistic model. The probit and cloglog predicted values both exhibit a similar sigmoid pattern when plotted against the logistic values, with higher predicted values than the logistic predictions for values less than about .35 and similar to lower values than logistic predictions for values greater than .4 and less than .9. The cloglog model shows more variance and a more distinct sigmoid curve pattern when plotted against the logistic predictions. Therefore, cloglog looks to estimate systematically larger and smaller probabilities depending at what range you are comparing witht he logistic predictions.

2.
```{r}
tree_parameters = tree.control(nobs=nrow(drug_use_train), minsize=10, mindev=1e-3)

weed_tree = tree(formula = recent_cannabis_use ~ ., data=drug_use_train ,control = tree_parameters)

summary(weed_tree)

```

```{r}
cv=cv.tree(weed_tree,FUN=prune.misclass)
best.cv = min(cv$size[which(cv$dev == min(cv$dev, na.rm = TRUE)) ])
cv$size
print(best.cv)
print('is the best')
```

```{r}
pruned_tree = prune.tree(tree = weed_tree, best = best.cv)
draw.tree(tree = pruned_tree, size=3, cex=1, nodeinfo=TRUE)
```

The country variable is the first split.


```{r}

pruned.predict.test = predict(pruned_tree, drug_use_test, type="class")

testtable = table(drug_use_test$recent_cannabis_use,pruned.predict.test)

testerror = 1-sum(diag(testtable))/sum(testtable)

testtable
true_postivies = testtable[2,2]
true_negatives = testtable[1,1]
false_postives = testtable[1,2]
false_negatives = testtable[2,1]
TPR = true_postivies / (false_postives+true_postivies)
FPR = false_postives / (true_negatives+false_postives)

print('TPR is')
TPR
print('FPR is')
FPR
```










