---
title: "Lab 1: Getting started with R and R markdown"
subtitle: "PSTAT 131/231, Spring 2018"
header-includes: \usepackage{float}
output: pdf_document
urlcolor: blue
---

> ### Learning Objectives
>
> - Complete installation of R and RStudio
> - First steps using RStudio
> - Getting help in R
> - Installing packages
> - Using R script files
> - Using Rmd files
> - Get to know markdown syntax
> - R code chunk options

------

## R and RStudio

- Make sure everybody has installed __R__
	- R for Mac: [https://cran.r-project.org/bin/macosx/](https://cran.r-project.org/bin/macosx/)
	- R for windows: [https://cran.r-project.org/bin/windows/base/](https://cran.r-project.org/bin/windows/base/)
- Make sure everybody has installed __RStudio__
	- RStudio download: [https://www.rstudio.com/products/rstudio/download/](https://www.rstudio.com/products/rstudio/download/)
- Difference between R-GUI and RStudio
	- R-GUI is a simply graphical user interface
	- RStudio is an _Integrated Development Environment_ (IDE)
		- It is much more than a simple GUI
		- It provides a nice working environment and development framework
- We are going to use mainly RStudio
- Keyboard shortcuts will increase your productivity significantly:
   
    <https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts>

-----

## More about RStudio

You will be working with RStudio a lot, and you will have time to learn
most of the bells and whistles RStudio provides. Think about RStudio as
your "workbench". Keep in mind that RStudio is NOT R. RStudio is an environment
that makes it easier to work with R, while handling taking care of many of
the little tasks than can be a hassle.


### A quick tour of RStudio

- Understand the __pane layout__ (i.e. windows) of RStudio
	- Source
	- Console
	- Environment, History, etc
	- Files, Plots, Packages, Help, Viewer
- Customize RStudio Appearance of source pane
    - font
    - size
    - background


### Using an R script file

Most of the times you won't be working directly on the console.
Instead, you will be typing your commands in some source file.
The basic type of source files are known as _R script files_.
Open a new script file in the _source_ pane, and rewrite the
previous commands.

You can copy the commands in your source file and paste them in the
console. But that's not very efficient. Find out how to run (execute)
the commands (in your source file) and pass them to the console pane.

### Getting help

Because we work with functions all the time, it's important to know certain
details about how to use them, what input(s) is required, and what is the
returned output.

There are several ways to get help.

If you know the name of a function you are interested in knowing more,
you can use the function `help()` and pass it the name of the function you
are looking for:

```r
# documentation about the 'abs' function
help(abs)

# documentation about the 'mean' function
help(mean)
```

Alternatively, you can use a shortcut using the question mark `?` followed
by the name of the function:
```r
# documentation about the 'abs' function
?abs

# documentation about the 'mean' function
?mean
```

- How to read the manual documentation
	- Title
	- Description
	- Usage of function
	- Arguments
	- Details
	- See Also
	- Examples!!!

`help()` only works if you know the name of the function your are looking for.
Sometimes, however, you don't know the name but you may know some keywords.
To look for related functions associated to a keyword, use double
`help.search()` or simply `??`
```r
# search for 'absolute'
help.search("absolute")

# alternatively you can also search like this:
??absolute
```
Notice the use of quotes surrounding the input name inside `help.search()`


### Installing Packages

R comes with a large set of functions and packages. A package is a collection
of functions that have been designed for a specific purpose. One of the great
advantages of R is that many analysts, scientists, programmers, and users
can create their own packages and make them available for everybody to use them.
R packages can be shared in different ways. The most common way to share a
package is to submit it to what is known as __CRAN__, the
_Comprehensive R Archive Network_.

You can install a package using the `install.packages()` function.
Just give it the name of a package, surrounded by quotes, and R will look for
it in CRAN, and if it finds it, R will download it to your computer.
```r
# installing
install.packages("knitr")
```

You can also install a bunch of packages at once:
```r
install.packages(c("readr", "ggplot2"))
install.packages(c("hflights", "tidyverse"))
```

Once you installed a package, you can start using its functions
by _loading_ the package with the function `library()`
```r
library(knitr)
```

### Your turn 1 

```{r, your-turn-1, eval=FALSE}
# Look for the man (i.e. help) documentation of the function `exp`

# Find out how to look for information about binary operators like `+` or `^`

```

-----

## Introduction to Markdown

Besides using R script files to write source code, you will be using other
type of source files known as _R markdown_ files. These files use a special
syntax called __markdown__.


### Get to know the `Rmd` files

In the menu bar of RStudio, click on __File__, then __New File__,
and choose __R Markdown__. Select the default option (Document),
and click __Ok__.

__Rmd__ files are a special type of file, referred to as a _dynamic document_,
that allows to combine narrative (text) with R code. Because you will
be turning in most homework assignments as `Rmd` files, it is important
that you quickly become familiar with this resource.

In order to __Knit PDF__, you need to install LaTeX distribution. If you try
this option without LaTeX, you will be greeted with the following error: 

```null
No TeX installation detected (TeX is required to create PDF output).
You should install a recommended TeX distribution for your platform:

  Windows: MiKTeX (Complete) - http://miktex.org/2.9/setup
  (NOTE: Be sure to download the Complete rather than Basic installation)

  Mac OS X: TexLive 2013 (Full) - http://tug.org/mactex/
  (NOTE: Download with Safari rather than Chrome _strongly_ recommended)

  Linux: Use system package manager
```

Locate the button __Knit HTML__ (the one with a knitting icon) and click on it
so you can see how `Rmd` files are rendered and displayed as HTML documents.

__Knit PDF__ is preferred for this class (much more flexible); however, if you
do not wish to install LaTeX system, you may __Knit HTML__ and print to pdf.

### Yet Another Syntax to Learn

R markdown (`Rmd`) files use [markdown](https://daringfireball.net/projects/markdown/)
as the main syntax to write content. It is a very lightweight type of
markup language, and it is relatively easy to learn.


### On your own time

Work through the markdown tutorial:
[www.markdown-tutorial.com](http://www.markdown-tutorial.com)

Your turn: After lab discussion, find some time to go through this
additional markdown tutorial
[www.markdowntutorial.com/](http://www.markdowntutorial.com/)

RStudio has a very comprehensive R Markdown tutorial:
[Rstudio markdown tutorial](http://rmarkdown.rstudio.com/)

-----

## R Markdown Code Chunks

The general format of an Rmd code chunk is as shown below

```
{r chunk_name, chunk_options}
```

### Option 1: `include=FALSE`

Here we just set up some knitr options we want the code to be 
_evaluated_ but **neither code or output to be** _displayed_:

```{r include_false_demo, include=FALSE}
library(knitr)
```

The code that was hidden in the final knit document was:

```
library(knitr)
```

### Option 2: `results='hide'`

Here we just set up some knitr options we want the code to be 
_evaluated_ but **the output not to be** _displayed_:

```{r results_hide_demo, results='hide'}
summary(cars)
```

### Option 3: `echo=FALSE`

Here we just set up some knitr options we want the code to be 
_evaluated_ but **only the output to be** _displayed_:

```{r echo_false_demo, echo=FALSE}
plot(cars$speed)
```

### Option 4: `eval=FALSE`

Here we just set up some knitr options we want the code to 
**not be evaluated** but **only the code to be** _displayed_:

```{r eval_false_demo, eval=FALSE}
min(cars$dist)
```

We didn't get the minimum value of the `distance` variable in `cars` 
dataframe displayed as expected here, just the R code command

### Option 5: `cache=TRUE`

Here if the code chunk remains unchanged between 'knitting', then it
won't be re-run in the next knit i.e. RStudio will cache (save) the output
from the previous knit and not re-run.

This is used to **speed up the R Markdown development process** by reducing
the time for each subsequent knit process.

If caching is not behaving in a manner you expect, consult this page to
troubleshoot: <https://yihui.name/knitr/demo/cache/>


```{r cache_true_demo, cache=TRUE}
max(cars$dist)
```

### Option 6: `warning=FALSE, message=FALSE`

Here `warnings` and `messages` of `R` output are suppressed. By default they
are not. In general you should not require this but there are some special 
cases where it is useful.

```{r warning_message_false_demo, warning=FALSE, message=FALSE}
library(knitr)
```

### Option 7: `Preferred for PSTAT 131/231`

In `PSTAT 131/231` we (i.e. Instructor and TAs) want to see your code and
output displayed in your submissions. So there should be no need to have
`echo=FALSE` or `results='hide'` or `eval=FALSE` (using `cache=TRUE` is up to
you):

```
{r chunk_name}
```
or
```
{r chunk_name, cache=TRUE}
```

### Other options

There are many other options for other functionality: e.g. plotting. You may
use them when appropriate. Here are some resources to help you:

- [Karl Borman's Rmd tutorial](http://kbroman.org/knitr_knutshell/pages/Rmarkdown.html)
- [Yihue Xie's Rmd Chunk Options page](http://yihui.name/knitr/options/)


### Your turn 2

```{r, your-turn-2, eval=TRUE}
# create the following variables
# x = seq(-1, 2, length.out=100)
# y = x^3 - 0.5*x^2 - x + 0.1

```

### Your turn 3

```{r, your-turn-3, eval=TRUE}
# plot x and y, and set chunk option to hide the plot command
# plot(x, y, type='o', pch=16, cex=0.5)
```

-----

Credit: Adopted from <https://github.com/ucb-stat133/stat133-fall-2016>, and
licensed under Creative Commons Attribution-NonCommercial-ShareAlike 4.0
International License.

